---
title: "Univariate TS - simulated data"
output: html_notebook
---

The objective is to become familiar with univariate time series (TS) models and tests within a controlled environment, utilizing simulated data. By controlling the environment, including knowledge of the true data-generating process, we can circumvent practical issues like outliers, discontinuities, and uninformative data. Instead, our focus is on analyzing the characteristics of TS models and developing a better understanding.       

## Packages
We will be using stats, ggplot2, ggpubr, gridExtra and forecast package (see also tseries package!). 

## Exercise 1: Simulate AR(1) Data

Simulate data from an AR(1) model. Do not use built-in functions. Program it from scratch. Next, in one figure, plot the series,  the autocorrelation (ACF) and the partial autocorrelation (PACF) functions (for lags 1 to 10). Next, vary the value of the autoregressive parameter ($\phi_1$) from the set {0, 0.5, 0.9, 1.0}. 

1. How does the pattern of the series change when $\phi_1$ ↑?
2. What happens to the ACF and PACF?
3. Can you determine, by inspecting the ACF and PACF, which time series model (AR, MA, ARMA) is appropriate for the simulated series?


```{r echo=TRUE}
# Load necessary packages
library(stats)        # package for statistics
library(ggplot2)      # package for plotting
library(ggpubr)       # for multiple figures in one
library(forecast)     # used for forecasting and some plots (ggAcf)

# Set seed
set.seed(1)

# Set burn-in observations (to wash out the effect of initial values) and the number of observations
nburn <- 100
nobs  <- 1000

# Set AR coefficient value
phi1 <- 0.5
truevar <- [??]

# Empty placeholder for the AR series
y <- numeric(nburn + nobs)

# Draw white-noise errors (N ~ (0, 1))
e <- rnorm(nburn + nobs)

# Simulate AR(1) process
for (i in 2:(nburn + nobs)) {
  y[i] <- [??]
}

# Drop burn-in observations 
y <- y[(nburn + 1):(nburn + nobs)]

# Change to time series data
y <- ts(y, start=1)


# # Plot series, ACF, and PACF
# # Open new window where plots will be shown
# dev.new()
# par(mfrow=c(1, 3))
# plot(y, type="l", main=sprintf("AR(1), phi1=%f", phi1))
# abline(h=0, col="red")
# acf(y, lag.max=10, main="ACF")
# pacf(y, lag.max=10, main="PACF")

# Create plot objects
plot_y  <- ggplot(data.frame(x = 1:nobs, y = as.numeric(y)), aes(x = x, y = y)) + 
  geom_line() + 
  labs(title = "AR(1)")+
  geom_hline(yintercept=0, color="red")
plot_ac  <- ggAcf(y,lag.max = 10) + labs(title="ACF")
plot_pac <- ggPacf(y, lag.max = 10) + labs(title="PACF")

# Display plot
ggarrange(plot_y,plot_ac,plot_pac,ncol = 3)

```

```{r}
# Display true and estimated mean and variance
cat("True mean is 0\n")
cat(sprintf("Sample mean is %f\n", mean(y)))
cat(sprintf("True variance is sig^2/(1-phi_1^2) = %f\n", truevar))
cat(sprintf("Sample variance is %f\n", var(y)))
```


## Exercise 2: Simulate MA(1) Data

Simulate data from an MA(1) model with $\theta_1$ values from the set {0, 0.5, 0.9, 1.0}. First, programe it from scratch. Next, utilize the stats package function for simulating data from ARIMA models.

```{r echo=TRUE}
# FROM SCRATCH

# Set seed
set.seed(1)

# Set burn-in observations (to wash out the effect of initial values) and the number of observations
nburn <- 100
nobs  <- 1000

# Set AR coefficient value
theta1 <- 0.5

# Empty placeholder for the AR series
y <- numeric(nburn + nobs)

# Draw white-noise errors (N ~ (0, 1))
e <- rnorm(nburn + nobs)

# Simulate AR(1) process
for (i in 2:(nburn + nobs)) {
  y[i] <- [??]
}

# Drop burn-in observations 
y <- y[(nburn + 1):(nburn + nobs)]

# Change to time series data
y <- ts(y, start=1)

# Create plot objects
plot_y  <- ggplot(data.frame(x = 1:nobs, y = as.numeric(y)), aes(x = x, y = y)) + 
  geom_line() + 
  labs(title = "MA(1)")+
  geom_hline(yintercept=0, color="red")
plot_ac  <- ggAcf(y,lag.max = 10) + labs(title="ACF")
plot_pac <- ggPacf(y, lag.max = 10) + labs(title="PACF")

# Display plot
ggarrange(plot_y,plot_ac,plot_pac,ncol = 3)

```


```{r}
# BUILT-IN FUNCTIONS

# set the number of observations
nobs <- 1000

# set MA parameter
theta1 <- c(0.5)

# simulate y_t
y <- arima.sim(list(order = c(0,0,1), ma = theta1), n = nobs, n.start = 100)

# Create plot objects
plot_y  <- ggplot(data.frame(x = 1:nobs, y = as.numeric(y)), aes(x = x, y = y)) + 
  geom_line() + 
  labs(title = "MA(1)")+
  geom_hline(yintercept=0, color="red")
plot_ac  <- ggAcf(y,lag.max = 10) + labs(title="ACF")
plot_pac <- ggPacf(y, lag.max = 10) + labs(title="PACF")

# Display plot
ggarrange(plot_y,plot_ac,plot_pac,ncol = 3)

```


## Exercise 3: Simulate ARMA(2,0) Data and Perform Statistical Tests

1. Simulate 200 data points from a stationary ARMA(2,0) model. Assume $\phi_1 = 0.6$ and $\phi_2 = 0.3$. Plot the data.
2. Test the data for stationarity using the Augmented Dickey-Fuller and Kwiatkowski-Phillips-Schmidt-Shin tests.
3. Inspect the autocorrelation (AC) and partial autocorrelation (PAC) functions. Select the number of autoregressive (AR) and moving average (MA) lags.
4. Estimate the true model (ARMA(2,0)).
5. Test the whiteness of the residuals using the Breusch-Pagan autocorrelation test.
6. Estimate ARMA(1,0) and ARMA(3,0) models, and discuss the differences between them.
7. Display information criteria for AR(1) to AR(3).

```{r}
# 3.1
# Seed
set.seed(1)
# set the number of observations
nobs <- 1000

# set AR parameters
phi <- c(0.5,0.2)

# simulate y_t
# ??arima.sim
y <- [??]

# plot
ggplot(data.frame(x = 1:nobs, y = as.numeric(y)), aes(x = x,y = y)) + 
         geom_line() + 
         labs(title = "AR(2)")+
         geom_hline(yintercept=0, color="red")
         

```

```{r}
# 3.2
# ADF test
library(urca)
# help(package="urca")
# What is the null hypothesis in the ADF test? H0: Series has a unit root! 

# NONE -  model without drift or time trend 
# MODEL: d_y_t = gamma*y_t-1 + e_t
# TEST:  t-stat for gamma, but compare to "special" critical values
# -      gamma or tau-statistic is the main focus 
# if [tau1 < CV_tau1] then we cannot reject gamma = 0 
y %>% ur.df(type ="none", lags = 5, selectlags = c("AIC")) %>% summary
```

```{r}
# 3.2
# ADF test
# DRIFT - model with drift
# MODEL: d_y_t = c + gamma*y_t-1 + e_t
# TEST:  gamma = 0, if true we have a unit root
# gamma or tau-statistic is the main focus 
# if [tau2 < CV_tau2] then we cannot reject gamma = 0
# if [phi1 < CV_phi1] then we cannot reject gamma = 0 AND (NOT OR) c = 0 
y %>% ur.df(type ="drift", lags = 5, selectlags = c("AIC")) %>% summary
```
```{r}
# 3.2
# ADF test
# TREND - model with trend
# MODEL: d_y_t = c + gamma*y_t-1 + a*t + e_t
# TEST:  gamma = 0, if true we have a unit root
# gamma or tau-statistic is the main focus 
# if [tau3 < CV_tau3] then we cannot reject gamma = 0
# if [phi2 < CV_phi2] then we cannot reject gamma = 0 AND (NOT OR) a = 0 
# if [phi3 < CV_phi3] then we cannot reject gamma = 0 AND a = 0 AND c = 0
y %>% ur.df(type ="trend", lags = 5, selectlags = c("AIC")) %>% summary
```
```{r}
# 3.2
# KPSS test
# What is the null hypothesis in the KPSS test? 
# Type: mu - difference stationary, tau - trend stationary
y %>% ur.kpss(type ="mu", lags = "short") %>% summary
```


```{r}
# 3.1 & 3.2
# Increase the persistence of the AR(2) model to phi1=.85, phi2=.1, so that phi_1 + phi_2 is < 1 but close to 1 (model is still stationary!). Simulate data and redo the tests. Set sample size to 100. Do the tests still give you the correct answer regarding stationarity?
# HELP:
# phi <- c(0.85,0.1)
# y <- arima.sim(list(order = c(2,0,0), ar = phi), n = nobs, n.start = 100)
```



```{r}
# Set back the original model: phi <- c(0.5,0.2), nobs <- 1000
 
# 3.3 Inspect AC and PAC functions,  select the model, AR&MA lags.
plot_ac  <- ggAcf(y,lag.max = 10) + labs(title="ACF")
plot_pac <- ggPacf(y, lag.max = 10) + labs(title="PACF")
# Display plot
ggarrange(plot_ac,plot_pac,ncol = 2)

# y %>% ggtsdisplay()
```


```{r}
# 3.4 Estimate the TRUE model. 
mdlEst <- Arima(y, order=c(2,0,0)) 

# interpret output
summary(mdlEst)
# https://stat.ethz.ch/R-manual/R-patched/library/stats/html/arima.html
```
```{r}
# perform statistical tests on the coefficients - interpret statistical significance
p_values = (1-pnorm(abs(mdlEst$coef)/sqrt(diag(mdlEst$var.coef))))*2 
p_values = round(p_values,2)
sprintf("P-values for the coefficients are: ")
p_values
```


```{r}
# 3.4 Test for whitness of the residuals by using an autocorrelation test.   
# interpret residual autocorrelation test
# what are the consequences of failing this test? 
checkresiduals(mdlEst,plot = TRUE, test = "LB")

# interpret Jarque-Bera test
# what are the consequences of failing this test? 
library(tseries)        # another package for modelling time series
jarque.bera.test(mdlEst$residuals)
```


```{r}
# 3.6 Estimate ARMA(3,0) although the true model is ARMA(2,0)
# the case of redundant regressor  
# discuss the consequences
mdlEst <- Arima(y, order=c(3,0,0)) 
summary(mdlEst)
p_values = (1-pnorm(abs(mdlEst$coef)/sqrt(diag(mdlEst$var.coef))))*2 
p_values = round(p_values,2)
sprintf("P-values for the coefficients are: ")
p_values
checkresiduals(mdlEst,plot = TRUE, test = "LB")
```

```{r}
# 3.6 Estimate ARMA(1,0) although the true model is ARMA(2,0)
# the case of missing an important regressor 
# discuss the consequences
mdlEst <- Arima(y, order=c(1,0,0)) 
summary(mdlEst)
p_values = (1-pnorm(abs(mdlEst$coef)/sqrt(diag(mdlEst$var.coef))))*2 
p_values = round(p_values,2)
sprintf("P-values for the coefficients are: ")
p_values
checkresiduals(mdlEst,plot = TRUE, test = "LB")
```


```{r}
# 3.7 Select the model by using information criteria
p_max = 5
AIC  = vector(length=p_max, mode="numeric")
AAIC = vector(length=p_max, mode="numeric")
BIC  = vector(length=p_max, mode="numeric")
for (p in 0:p_max) {
  # estimate the model
  mdlEst <- Arima(y, order=c(p,0,0))
  # store information criteria
  AIC[p+1]  <- mdlEst$aic
  AAIC[p+1] <- ifelse(!is.null(mdlEst$aaic), mdlEst$aaic, NA) 
  BIC[p+1]  <- mdlEst$bic
} 
# AIC <- lapply(seq(0,p_max,1), \(p) {Arima(y, order=c(p,0,0))$aic})
# AIC <- sapply(seq(0,p_max,1), \(p) {Arima(y, order=c(p,0,0))$aic})
# rm(list = c("AIC", "AAIC", "BIC"))
p_aic <- which.min(AIC) - 1 
p_bic <- which.min(BIC) - 1 
sprintf("Optimal lag according to AIC is %d",p_aic)
sprintf("Optimal lag according to BIC is %d",p_bic)
```
